Staged typed type-preserving tagless interpreters, \emph{finally}

### 2007-03-24 matrix/tagless-final.tex
misc corrections to the text

\title{Title Text}
\subtitle{Subtitle Text, if any}

@@

@@

The first attempt at the interpreter: take the object language term
such as test1 and give us the value
\begin{code}
  let rec eval0 env = function 
  | V VZ -> List.hd env
  | V (VS v) -> eval0 (List.tl env) (V v)
  | B b -> b 
  | L e -> fun x -> eval0 (x::env) e
  | A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

If our meta-language (the language in which we wrote test1
and that interpreter) were untyped, the above code would be acceptable.
The |L e| line exhibits interpretative overhead (see descr in Walid02
paper): |eval0 env' e| will be executed every time the result of
evaluating |L e| is applied. Staging can be used to remove the
interpretative overhead. Cite some paper -- see Walid02, Sec 1.1 and
1.2. They say it well.

@@

We can then write the interpreter, and it works
\begin{code}
  let rec eval env = function
  | V VZ -> List.hd env
  | V (VS v) -> eval (List.tl env) (V v)
  | B b -> UB b
  | L e -> UA (fun x -> eval (x::env) e)
  | A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
\end{code}
whose inferred type is |u list -> exp -> u|,

@@

Unfortunately, the result is tagged: |UB true| rather than just
|true|. We also note that the eval function is partial in two
respects: first there is an inexhaustive pattern-matching in the 
line |A (e1,e2)|.
Second, the occurrences of |List.hd| and |List.tl|.
The failure of the latter corresponds to the evaluation of `open'
terms. One can see that if all our terms are closed, List.hd and
List.tl above will never be applied to the empty list. The partial
pattern match in the |A (e1,e2)| line raises an exception when we try to
evaluate something like |A (B true, B false)|, that is, the first
operand is not a function.

Suppose our object language were typed, too. Then terms like
|A (B true, B false)| would be ill-typed. Thus, if our evaluator
accepted only typed terms, then the missing alternatives in the |A|
rule would not occur. Alas, what is obvious to us is not obvious to
the meta-language. There is still the pattern match, and we still have
to attach and eliminate the UB and UA tags, even when there is no
longer a need for them. See Walid02, Sec 1.4 for more detail.

Of course there is also the issue how to design a type system to type
terms like test1 (GADT seem necessary). Also, how to design the
typesystem to keep track of open/closed status of terms, so we can
statically assure that only closed terms are passed to eval and that
the Nil checking in List.hd and List.tl can be eliminated as well. The
latter is even more interesting problem requiring even more advanced
type systems (cite env classifiers, Nanevski, Rowan Davies).

@@

Now, let us define our object language (embedded DSL) in a different
way. Instead of using algebraic datatypes to encode the terms of the
language in the meta-language (reminiscent to the initial approach),
we will be using functions as term constructors (final approach):

@@

That is all of the interpreter. The previous term |test1| now reads
\begin{code}
  let testf1 = app (lam varZ) (b true)
\end{code}

To evaluate the term, we merely need to apply that to the empty
environment

@@

Now we see that the result has no tags. The above functions |b|,
|varZ|, |app| constitute the evaluator. It is patent that the
interpreter has no tags and no pattern matching. A term |b true| is
evaluated to a boolean and a term |lam varZ| is evaluated to a
function, with no tags attached. The application rule than applies that
function without any need for pattern-matching. The evaluator has
another feature: evaluation of open terms leads to a type error rather
than a run-time error. Indeed, the following gives a type error

\begin{code}
  let testf1bad = app (lam (varS varZ)) (b true)
  let testf1badr = testf1bad ()
\end{code}
that the initial environment should be a tuple rather than () --
meaning the term requires non-empty initial environment, the term is
open.

Thus all the errors that previously reported at run-time (open terms,
applications of non-functions) are now reported at compile
time. Running of the interpreter assuredly gives no run-time errors
because its code uses no run-time error-raising constructs.

The interpreter uses no Universal type or pattern matching, and it
is obvious to the compiler that non-termination can only result from
interpreting fix.  If the source language is strongly normalizing, then
our interpreter is total.

@@

@@

Introduce the language, 
  The language is simply-typed lambda-calculus with fixpoint,
  integers, booleans and comparison.

\begin{code}
  Lam hoas_fn | App e e | Fix hoas_fn |
  I Int | B Bool | Add ie1 ie2 | Mul ie1 ie2 | Leq ie1 ie2 |
  IF b e-then e-else
\end{code}

this time with HOAS rather than deBruijn indices. The language is close 
to the language of Xi03, but without the polymorphic lift and more 
constants so we can write better examples, enough to express
Fibonacci, factorial, and power.

@@

The previous test now reads |app (lam (\x -> x)) (bool True)| and the
inferred type is |Symantics repr => repr Bool|.

@@

\begin{code}
testpowfix () = lam (\x ->
                      fix (\self -> lam (\n ->
                        if_ (leq n (int 0)) (int 1)
                            (mul x (app self (add n (int (-1))))))))
testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
\end{code}

OCaml: functors (give the signature of
Symantics without the 'sv part)

@@

\begin{code}
module EX(S: Symantics) = struct
 open S
 let testpowfix () = lam (fun x ->
                      fix (fun self -> lam (fun n ->
                        if_ (leq n (int 0)) (fun () -> int 1)
                            (fun () -> mul x (app self (add n (int (-1))))))))

 let testpowfix7 = 
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
The unit in |testpowfix| is for the sake of value restrcition (which
interferes with the environment classifier inference).

Thus the DSL expression has the type of the functor, from |Symantics|
to the value.

@@

the sv type problem: the inductive type function 'dv -> 'sv
its GADT solution. 
its type-level function solution: incope1.hs 
  Need to thank Martin Sulzmann

@@

>>>>

\title{Staged typed type-preserving tagless interpreters, \emph{finally}}
\subtitle{Subtitle Text, if any}

@@

We have programmed out interpreters in OCaml/MetaOCaml (cite
www.metaocaml.org) and Haskell. The complete code is available 
% connot give URL in the ICFP paper due to double-blind reviews
in the supplemental material to the paper. We will use both
(Meta)OCaml and Haskell for our examples throughout the paper.

@@

The first attempt at the interpreter, the function eval. It takes the
object language term such as test1 and gives us its value.
The first argument to |eval| is the environment, initially empty,
which is the sequence of values associated with free variables in the
interpreted code.
\begin{code}
  let rec eval0 env = function 
  | V VZ -> List.hd env
  | V (VS v) -> eval0 (List.tl env) (V v)
  | B b -> b 
  | L e -> fun x -> eval0 (x::env) e
  | A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

Suppose our our meta-language (the language in which we wrote test1
and that interpreter) were untyped. The above code would be acceptable.
The |L e| line exhibits interpretative overhead (see descr in Walid02
paper): |eval0 env' e| will be executed every time the result of
evaluating |L e| is applied. Staging can be used to remove the
interpretative overhead. Cite some paper -- see Walid02, Sec 1.1 and
1.2. They say it well.

@@

We can then write the interpreter, which is accepted by the typecheker
\begin{code}
  let rec eval env = function
  | V VZ -> List.hd env
  | V (VS v) -> eval (List.tl env) (V v)
  | B b -> UB b
  | L e -> UA (fun x -> eval (x::env) e)
  | A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
\end{code}
and whose inferred type is |u list -> exp -> u|. Now we can evaluate
our sample |test1| term:

@@

Unfortunately, the result is tagged: |UB true| rather than just
|true|. We also note that the eval function is partial in two
respects: first there is an inexhaustive pattern-matching in the 
line |A (e1,e2)|.
Second, the occurrences of |List.hd| and |List.tl|.
The failure of the latter corresponds to the evaluation of `open'
terms. One can see that if all our terms are closed, |List.hd| and
|List.tl| above will never be applied to the empty list. The partial
pattern match in the |A (e1,e2)| line raises an exception when we try to
evaluate something like |A (B true, B false)|, that is, the first
operand is not a function.

Suppose our object language were typed, too. Then terms like
|A (B true, B false)| would be ill-typed. Thus, if our evaluator
accepted only typed terms, then the missing alternatives in the |A|
line would not occur. Alas, what is obvious to us is not obvious to
the meta-language. There is still the pattern match, and we still have
to attach and eliminate the UB and UA tags, even when there is no
longer a need for them. See Walid02, Sec 1.4 for more detail.

Of course there is also the issue how to design a type system to type
terms like test1 (GADT seem necessary). Also, how to design the
typesystem to keep track of open/closed status of terms, so we can
statically assure that only closed terms are passed to eval and that
the empty list checking in |List.hd| and |List.tl| can be eliminated
as well. The latter is even more interesting problem requiring even
more advanced type systems (cite env classifiers, Nanevski, Rowan
Davies).

@@

Now, let us define our object language (embedded DSL) in a different
way. Instead of using algebraic datatypes to encode the terms of the
language in the meta-language (reminiscent to the initial approach),
we will be using functions as term constructors (the final approach):

@@

That is all of the interpreter. Our sample term now reads
\begin{code}
  let testf1 = app (lam varZ) (b true)
\end{code}
which is almost the same as before, only written with lower-case
identifiers. To evaluate the term, we merely need to apply 
the term to the empty environment

@@

The result now has no tags. The above functions |b|,
|varZ|, |app| constitute the evaluator. It is patent that the
interpreter has no tags and no pattern matching. The term |b true| is
evaluated to a boolean and a term |lam varZ| is evaluated to a
function, with no tags attached. The application rule applies that
function without any need for pattern-matching. The evaluator has
another feature: evaluation of open terms leads to a type error rather
than a run-time error. Indeed, the following gives a type error

\begin{code}
  let testf1bad = app (lam (varS varZ)) (b true)
  let testf1badr = testf1bad ()
\end{code}
that the initial environment should be a tuple rather than () --
meaning the term requires non-empty initial environment: the term is
open.

Thus all the errors that were previously reported at run-time (open terms,
applications of non-functions) are now reported at compile
time. Running of the interpreter assuredly gives no run-time errors
because its code uses no operations that could possibly raise a
run-time error.

The interpreter uses no Universal type or pattern matching, and it
is obvious to the compiler that non-termination can only result from
interpreting |fix|.  If the source language is strongly normalizing, then
our interpreter is total.

@@

The structure of our paper is as follows. Section XXX ...
We finally conclude.

@@

Introduce the language, The language is simply-typed lambda-calculus
with fixpoint, integers, booleans and comparison, whose BNF grammar is

\begin{code}
  e ::= Lam hoas_fn | App e e | Fix hoas_fn |
  I int_literal | B bool_literal | Add e1 e2 | Mul e1 e2 | Leq e1 e2 |
  IF b e-then e-else
\end{code}

This time we use the higher-order abstract syntax (cite: but not
Elliott. Cite Dale Miller!) rather than deBruijn indices.  The
language is close to the language of Xi03, without the polymorphic
lift but with more constants so we can write better examples, enough to
express Fibonacci, factorial, and power.

@@

Our sample code test1 now reads |app (lam (\x -> x)) (bool True)|,
whose inferred type is |Symantics repr => repr Bool|.

@@

\begin{code}
testpowfix () = 
  lam (\x -> fix (\self -> lam (\n ->
        if_ (leq n (int 0)) (int 1)
            (mul x (app self (add n (int (-1))))))))
testpowfix7 () = 
  lam (\x -> app (app (testpowfix ()) x) (int 7))
\end{code}
The dummy unit |()| argument above is to avoid the monomorphism
restriction and so that the type of |testpowfix| and |testpowfix7|
would remain polymorphic in |repr|. Instead of supplying the dummy
argument, we could have given the terms explicit polymorphic
sugnature. We however prefer if the typechecker of the meta-language
(Haskell) inferred the types of the object terms for us. We could also
avoid the |()| argument by switching off the monomophism restriction
with a compiler flag.

OCaml: functors.

@@

The power example and our sample term |test1| are encoded as
\begin{code}
module EX(S: Symantics) = struct
 open S
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = 
   lam (fun x ->fix (fun self -> lam (fun n ->
         if_ (leq n (int 0)) (fun () -> int 1)
             (fun () -> mul x (app self (add n (int (-1))))))))

 let testpowfix7 = 
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
The dummy unit argument in |testpowfix| is the artefact of MetaOCaml
(which we will be using to compile our tests in Sections below). To be
more precise, the environment qualifier (which appears as the |'c|
type parameter in the signature of Symantic functions) must remain
uninstantiated in order for us to `run' the code. That means, the type
must be polymorhic. The value restriction then dictates 
the definitions of our object terms must look syntactically like
values. Alternatively to giving the |()| argument, we could have used
rank-2 record types of OCaml to maintain the necessary polymorphism.

Thus the object expressions (expressions of the embedded DSL) have in
OCaml the type of the functor, from |Symantics| to the value. This
essentially the same as |Symantics repr => | constraint in the case of
the embedding of the DSL in Haskell.

@@

the sv type problem: the inductive type function 'dv -> 'sv
its GADT solution. 
its type-level function solution: incope1.hs

@@

\subsection*{Acknowledgments}
We woul like to tank Martin Sulzmann for helpful discussions.
